18.821 Project 2

### 2014-10-27 paper/paper.tex
added the Young tableaux stuff

% Special symbols, etc.
\usepackage{amssymb,amsbsy,latexsym}
\usepackage{amsmath}
\usepackage{graphics, subfigure, float} 
\usepackage{cancel}
%\usepackage{todonotes}
% Encoding settings
\usepackage[latin1]{inputenc}
\usepackage[american]{babel}
\usepackage[T1]{fontenc} 
\usepackage{tikz}

@@

>>>>

% Special symbols, etc.
\usepackage{amssymb,amsbsy,latexsym,ytableau}
\usepackage{amsmath}
\usepackage{graphics, subfigure, float} 
\usepackage{cancel}
%\usepackage{todonotes}
% Encoding settings
\usepackage[latin1]{inputenc}
\usepackage[american]{babel}
\usepackage[T1]{fontenc} 
\usepackage{tikz}

@@

\section{Terms in the sequence $s_n(321)$}

Unfortunately, there is no clear way to reformulate Lemma 2 for the case of avoiding the permutation 321 or 123. As a consequence, the proof in the previous section simply does not hold. Fortunately, however, we still have a closed form expression for $s_n(321)$ (recall that $s_n(321)=s_n(123)$ by our flipping principle). Surprisingly, in fact, the sequence remains the same: $s_n(321)=C_n$, the $n^{th}$ Catalan number. In order to prove this result, however, we will need the machinery of Young tableaux. What follows is a brief exposition of the key results, adapted from Stanley's \emph{Algebraic Combinatorics}.

\begin{definition}
A Young diagram of a partition $\lambda=\{\lambda_1, \lambda_2, \ldots, \lambda_n\}$ of the integer $\sum_{i=1}^n \lambda_i$ is a left-justified array of squares, with $\lambda_i$ squares in the $i$th row.
\end{definition}
\begin{example}
The Young diagram of $(4, 3, 1, 1)$ looks like:
\[
\ytableausetup{centertableaux}
\begin{ytableau}
\ & \ & \ & \ \\
\ & \ & \ \\
\ \\
\
\end{ytableau}
\]
\end{example}

\begin{definition}
A standard Young tableau (or SYT) consists of the Young diagram $D$ of some partition $\lambda$ of an integer $n$, together with the numbers $1, 2, \ldots, n$ inserted into the squares of $D$, so that each number appears exactly once, and every row and column is \emph{increasing}. We call $\lambda$ the \emph{shape} of the SYT.
\end{definition}

\begin{example}
There are five SYT of the shape $(2, 2, 1)$. They are given by
\[\ytableausetup{aligntableaux=top}
\begin{ytableau}
1 & 2 \\
3 & 4\\
5
\end{ytableau}\ \ \ \ \
\hfill
\begin{ytableau}
1 & 2 \\
3 & 5\\
4
\end{ytableau}\ \ \ \ \
\hfill
\begin{ytableau}
1 & 3 \\
2 & 4\\
5
\end{ytableau}\ \ \ \ \
\hfill
\begin{ytableau}
1 & 3 \\
2 & 5\\
4
\end{ytableau}\ \ \ \ \
\hfill
\begin{ytableau}
1 & 4 \\
2 & 5\\
3
\end{ytableau}.\]
\end{example}

\begin{definition}
Let $u$ be a square of the Young diagram of the partition $\lambda$. Then the hook $H(u)$ is the set of all squares directly to the right of $u$ or directly below $u$, including $u$ itself. The size of $H(u)$ is called the hook length of $u$, and is denoted $h(u)$.
\end{definition}

\begin{example}
In the Young diagram of the partition $(4, 2, 2)$ below, each square $u$ contains its hook length $h(u)$.
\[\ytableausetup{centertableaux}
\begin{ytableau}
6 & 5 & 2 & 1\\
3 & 2\\
2 & 1
\end{ytableau}\]
\end{example}

\begin{theorem}
Let $\lambda$ be a partition of $n$. Then $f^\lambda$, the number of SYT of shape $\lambda$, is given by
\[f^\lambda=\frac{n!}{\prod_{u\in \lambda} h(u)},\]
where the notation $u\in \lambda$ means that $u$ ranges over all squares of the Young diagram of $\lambda$.
\end{theorem}

The theorem above is called the \emph{hook-length formula}.

\begin{example}
The diagram above for the hook lengths of $\lambda=(4, 2, 2)$ tells us that the number of SYT of shape $\lambda$ is given by
\[\frac{8!}{6\cdot 5\cdot 2 \cdot 1 \cdot 3 \cdot 2 \cdot 2\cdot 1}=56.\]
\end{example}

It is an easy application of the hook-length formula to prove the following theorem:

\begin{theorem}
The number of SYT of shape $(n, n)$ is given by $C_n$, the $n^{th}$ Catalan number.
\end{theorem}

We need only one more result before we are ready to prove our central result.

\begin{theorem}
There exists an algorithm (called the \emph{RSK algorithm}) which defines a bijection between the set of permutations in $S(n)$ avoiding a decreasing subsequence of length 3 and the set of pairs of size-$n$ SYT of the same shape and at most two rows. 
\end{theorem}
Using all this machinery and all these results, we are finally able to prove the closed-form expression for $s_n(321)$.
\begin{theorem}
The total number of permutations of $\{1,2,3,\ldots, n\}$ that avoid the order 321 as a subsequence is $C_n$, where $C_n$ is the $n^{th}$ Catalan number.
\end{theorem}
\begin{proof}
By the first part of Theorem 12, there are $s_n(321)$ elements in the set $A$ of pairs of size-$n$ SYT of the same shape and at most two rows. By Theorem 11, the set $B_n$ of SYT of shape $(n, n)$ has its size given by $C_n$. We will construct a bijection between the sets $A_n$ and $B_n$: because $|A_n|=s_n(321)$ and $|B_n|=C_n$, it will therefore follow that $s_n(321)=C_n$, as desired.

Given any pair of size $n$ Young tableaux of the same shape and at most two parts, take the second tableau and invert its numbers: that is, send $i$ to $2n+1-i$ for all $i$. Now rotate the Young tableau and `fit' it into the first tableau. We demonstrate this process for a pair of Young tableaux with $n=6$: we begin with
\[\ytableausetup{aligntableaux=top}
\begin{ytableau}
1 & 2 &3&4\\
5 & 6
\end{ytableau} \ \ \ \ \ 
\hfill
\begin{ytableau}
1 & 3 & 4 & 5\\
2 & 6
\end{ytableau}.\]
Inverting the numbers of the second tableau gives us
\[\ytableausetup{aligntableaux=top}
\begin{ytableau}
1 & 2 &3&4\\
5 & 6
\end{ytableau} \ \ \ \ \ 
\hfill
\begin{ytableau}
12 & 10 & 9 & 8\\
11 & 7
\end{ytableau}.\]
Finally, rotating the second Young tableau and fitting it into the first gives us
\[\ytableausetup{aligntableaux=top}
\begin{ytableau}
1 & 2 &3&4&7&11\\
5 & 6&8&9&10&12
\end{ytableau},\]
a valid Young tableau of shape $(6, 6)$.

Beginning with a pair of size $n$ Young tableaux of shape $\lambda$ and at most two parts, it is easy to see that this process always yields a valid Young tableau of shape $(n, n)$. Indeed, we can also define the inverse: given any Young tableau for the partition $(n, n)$, we can find the shape formed by all numbers greater than $n$: split it off from the original tableau, rotate it, and invert the numbers to get a pair of size $n$ Young tableau of the same shape and at most two parts. These processes can be easily checked to be inverses: hence the two sets are of the same size, and the proof follows.
\end{proof}
